Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-09 analysis/paper/paper.Rmd
editting based on comments

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

```

@@

Taiwan provides an ideal case study of indirect effects of colonialism because it has archaeological evidence spanning its initial involvement in regional exchange networks of East Asia and Southeast Asia since the \tenth century. It became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia [@Liu2017]. Despite colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 study \the long-term record of Heping Dao\ in northern Taiwan and explore the effects of European colonization on the local populations by focusing on the environmental change and introduction of diseases, with a little mention of indigenous social complexity. Yilan in Northeastern Taiwan is an especially unique location for exploring indirect effects due to its physically isolated location. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which indigenous people in Yilan could interact with other groups from different regions, including other native Indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into Indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on Indigenous populations can be detected. I focus on  the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems, and encountered \several waves of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on Indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan due to overall population declines and acculturation\.

@@

Prior to European colonization, the Indigenous people in northern Taiwan had already established their own exchange network that has been described as “inter-insular trade” by Chen @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan was increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Due to its location on the periphery of major trade ports, \Keelung and \Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. A indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used as to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch VOC defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend annual ceremony for demonstrate their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from the year of 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several Indigenous groups and found that golden necklaces were not used in Yilan at that time, but elderly people still used beads.

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called “subli”, by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as a heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. It was usually the female shamans conducting divination rituals that passed the beads down from their ancestors. Despite most Kavalan  people today not knowing the specific origin of their agate beads, the value of beads is still highlighted through oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

@@

```{r, KWL-map, fig.cap = "Map showing the relative location of Kiwulan, Keelung, and Tamsui in northeastern Taiwan", out.width= "400px", echo=FALSE}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 squares (4 m by 4 m) reaching 3,814 m2 in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of excavation (\Figure 2). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age, and historical period starting from the 17th century with European contact in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5, L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The start of European contact in Kiwalan is indicated in the archaeological record by the presence of An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. I used the presence and absence of An-ping jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recongnise this as the Chinese colonization period, which is archaeological visible at Kiwulan by \Chinese porcelains and bricks that brought by Chinsese communities [@Hsieh2009]. Also, the description of Chinese migration can be identified from Chinese official records [@Chen1963].

@@

ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Contact",
    `6-layers` == 4 ~ "European Contact",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Contact", 
                                 "Chinese Contact"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))

sample_size <- nrow(ornaments_period)
total_AD <- nrow(kwl_upper)
sample_pro <- round(sample_size/total_AD, 1)
```

```{r read-in-the-spatial-data, eval = TRUE}
library(sf)
library(cartography)
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
```

```{r tidy-spatial-data, eval = TRUE}
# 40 sampling units
sample_units <-
  AD_data %>% 
  semi_join(units_40, by = c (`坑號` = "the_sq"))

# join two dataset to get coordinate
ornaments_period_join <-
  ornaments_period %>%
  mutate(period = as.factor(period)) %>% # seems need this line first 
  group_by(Pit_No, period) %>% 
  count(Categories) %>% 
  inner_join(AD_data, by = c("Pit_No" = "坑號"))
```

```{r, KWL-sam-area, fig.cap = "Forty squares were sampled from the largest open area (in red with square ID number)", out.width= "70%", echo=FALSE}

# Need to say that this is a map of excavation areas at Kiwulan, because the reader cannot see the site name anywhere in this figure. Need to add in here that each square is 4 x 4 m, and say which way north is. Or add to the map and scale and north arrow.

# mapping sampling area
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, alpha = 0.9) +
  theme_minimal() +
  theme_void() 
```

@@

In general, complete ornaments with clear context were mostly found in burials that helps to understand how people wear them. In addition to burials, large amount of ornaments were found at household area that were identified by a series of post holes across the space. Here I focus on the ornaments found at post-hole area considering the overall quantity and the comparison across the space. Ornaments were made from a wide variety of raw materials, in which glass, agate and copper are the common materials that were thought to be introduced through trading with the Europeans and Chinese traders since no clear evidence of waste by-products in the archaeological record [@Chen2007]. Although a possibility of small workshops that individuals were able to produce golden beads from prepared glass materials by heating [@Cheng2008], no evidence of accompanying tools or materials in proportion to the large amount of golden beads were found. The shape of ornaments include beads, bracelets, rings, bells, pendants, and knitted objects. It shows that the shape is closely related to what kind of raw material it was made from. For example, beads were only made of glass and agate, while copper ornaments were in the form such as bells, bracelets, and rings [@Chen2007].

# Samples and Methods

```{r plot-all-ornaments-count, fig.cap= "Frequency of ornaments", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

knitr::kable(ornaments_shape,
             caption = "A table of ornament subtypes")
```

@@

Ornaments from the post-hole area accounts for the \majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts \for `r sample_pro` % of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in each major category is shown in Figure \@ref(fig:ornaments-image).

<---include the number of ornaments from burials above--->

<---Something sounds wrong here. You say ‘majority’ but then later in the paragraph you say it’s 0.5% of the total number? How can 0.5% be the majority? --->

```{r, plot-counts-by-period, fig.cap = "Frequency of each category of ornament with a separate plot showing different time periods individually"}
### delete one plot

@@

# histogram facetting for three periods
orna_periods_facet <-
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# stack histogram
orna_all_viridis <-
ggplot(ornaments_count) +
  geom_col(aes(Categories, n, fill = period)) +
  labs(y = "Frequency", x = "Categories") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  scale_fill_viridis_d()

library(cowplot)
library(viridis)
# add two plots together
ggdraw() +
  draw_plot(orna_all_viridis + 
            theme(legend.justification = "top"), 0, 0, 1, 1) +
  draw_plot(orna_periods_facet + scale_color_viridis(discrete = TRUE) + 
              theme(legend.justification = "bottom"), 0.5, 0.15, 0.5, 0.5) +
  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 0.7), size = 15)
```

@@

In this section I report the tests of the three hypotheses about indirect impacts of European colonization on the Indigenous community at Kiwulan. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

```{r scatter-plot-bead-by-perio, eval= FALSE}
# plot for intact golden bead by period
ornaments_intact_gold <-
  ornaments_period %>% 
  filter(Condition %in% c('complete', 'near-complete'),
         Categories == 'Golden bead')

ggplot(ornaments_intact_gold) +
  geom_point(aes(Length, Width, col = Perforation_average)) +
  facet_wrap(~ period) +
  labs(x = "Length (mm)",
       y = "Width (mm)",
       title = "Golden beads") +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))

# plot for intact glass bead by period
# need to deal with the original data

# plot for intact agate bead by period
ornaments_intact_agate <-
  ornaments_period %>% 
  filter(Condition %in% c('complete','near-complete'),
         Categories == 'Agate bead')

ggplot(ornaments_intact_agate, 
       aes(Length, Width, col = period))+
  geom_point() +
  labs(x = "Length (mm)",
       y = "Width (mm)",
       title = "Agate beads") +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

@@

```{r scatter-plot-metal-by-period, eval= FALSE}
# plot for intact bell by period
ornaments_intact_bell <-
  ornaments_period %>% 
  filter(Condition %in% c('complete', 'near-complete'),
         Categories == 'Bell',
         Type != 'unclassified') %>% 
  mutate(Length = as.numeric(gsub("\\+", "", `Length(mm)`))) #remove +

>>>>

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 study both archaeological and historical records of Heping Dao in northern Taiwan that was the location of the colony of San Salvador founded by the Spanish in 1626 and took over by the Dutch in 1642. They uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. It shows a direct impact on the local indigenous populations by disturbing their long settlement pattern and changing local landscape. Different from the view of a Spanish fort, Yilan in Northeastern Taiwan is an especially unique location for exploring indirect colonial effects on local indigenous settlements due to its physically isolated location. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which indigenous people in Yilan could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since tenth century, and encountered large waves of Han Chinese immigration after 17th century since Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007, ch. 11]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community.

@@

Northeastern Taiwan was part of the local exchange network in northern Taiwan, where the indigenous people had already established their own exchange network before European arrived that has been described as “inter-insular trade” by Chen @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. The exchange network in northern Taiwan was increasingly influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on the shipping routes [@Chen2005]. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007]. Due to its location on the periphery of major trade ports, Keelung and Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for a religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. An indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery, and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch VOC defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from the year of 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan had the custom to wear ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called “subli”, by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. It was usually the female shamans conducting divination rituals that passed the beads down from their ancestors. Despite most Kavalan people today not knowing the specific origin of their agate beads, the value of beads is still highlighted through oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

@@

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text", out.width= "400px", echo=FALSE}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Contact",
    `6-layers` == 4 ~ "European Contact",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Contact", 
                                 "Chinese Contact"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))
```

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", out.width= "70%", echo=FALSE}

# add to the map and scale and north arrow. 
knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age to the historical period starting from the 17th century with European contact in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. I take approximate indicators of the start of European contact at Kiwulan as the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages, and the jar shapes found at Kiwulan are typical of those found elsewhere in V.O.C sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recognize this as the Chinese contact period, which is archaeologically visible at Kiwulan by a large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration can be seen from the Chinese official records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan from sampling area", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

# making table
knitr::kable(ornaments_shape,
             caption = "A table of ornament subtypes")

# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

ornament_context <-
  data.frame("contexts" = c("post hole area", "burials", "middens"),
           "sampling" = c(406, 3173, 27),
           "total_KWL" = c(873, 6769, 55))

ornament_context %>% 
  mutate(percent_context = round(sampling/sum(ornament_context$sampling)*100, 1),
         sam_total_percent = round(sampling/total_KWL*100, 1)) %>% 
  add_row(contexts = "sum", 
          sampling = sum(ornament_context$sampling), 
          total_KWL = sum(ornament_context$total_KWL))
```

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole area, 3173 pieces from burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments because most beads were preserved well in the form of bead strings that consist of hundreds to thousand pieces. In this paper I focus on the post hole area where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole area represents Kiwulan domestic space that provides clear contexts and basis for comparison to see if there is any spatial pattern of distribution. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in each major category is shown in Figure \@ref(fig:ornaments-image).

Ornaments were found in 30 burials from the sampling area as grave goods. Their positions in burials in association with particular parts of human remains indicate the purpose of body adornment. For example, agate beads, golden beads, and small metal rings were often found near skulls that indicate the use as necklaces or earrings, while big metal rings were found wearing on radius and ulna bones as bangles. This pattern corresponds to the texts regarding the body decorations of indigenous culture from Chinese historical documents [@Chen1963; @Ke1993]. Agate beads were mostly found in female burials with an average of 4 to 5 pieces, while golden beads were found in both genders with an average of 2 to 3 pieces.

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods."}
### delete one plot

@@

# histogram facetting for three periods
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "categories", y = "Frequency") +
  theme_minimal(base_size = 12)
```

@@

In this section I report the tests of the three hypotheses about indirect impacts of European colonization on the indigenous community at Kiwulan. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

@@
